---
title: "STAT 345 Final Project - What's Your Address?"
author: "STAT 345"
output: html_document
---

1.  *Maps* The US Census Bureau provides mapping shape files for a variety of variables across the entire United States. You can explore these at <https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html>. Make a map that displays county boundaries for the entire United States. You may use whichever package for plotting that you'd like, though you might find the 'ggmap' and 'rdgal' packages (and related tutorials) informative here.

2.  *Roads* Your goal in this step is to summarize (tabulate) the road suffixes for each county in the United States. County-level data can be found at <https://www2.census.gov/geo/tiger/TIGER2023/ROADS/>. There are over 3000 files here, with file names that include the 5-digit GEOID for the county. As you process this data, you'll need to manage your memory in R. You'll likely want to download the file, unzip it, read it, summarize it, and then remove the file from memory. This is a great place for a function! Be sure to include the GEOID in your summary data for the next step.

3.  *Putting it Together* Merge the county shape files with your county-level summaries. The main plot to create for the project is to color your map from step 1 based on the most common road suffix (summary from step 2). Beyond this plot, feel free to be creative when making at least two additional plots.

```{r}
# Libraries
library(sf)
library(dplyr)
library(stringr)
library(purrr)
```

```{r}
# Direct download of the county shapefile (no need to loop - one file covers all US counties)
url <- "https://www2.census.gov/geo/tiger/TIGER2023/COUNTY/tl_2023_us_county.zip"
download.file(url, destfile = "county.zip")
unzip("county.zip", exdir = "county_shape") #unzip this to folder called "county_shape"

# Reading in the shapefile
counties <- st_read("county_shape/tl_2023_us_county.shp") #reading shapefile into R as a spatial df
geoids <- counties$GEOID #extracting GEOID column which need later in function
```

```{r}
# This function takes a county GEOID, downloads the roads shapefile, and returns the most common road suffix for that county. Repeating process for all 3000+ counties.
process_county_roads <- function(geoid) {
  tryCatch({
    
    # 1. Download the zip file for this county
    url <- paste0("https://www2.census.gov/geo/tiger/TIGER2023/ROADS/tl_2023_", geoid, "_roads.zip")
    tmp_zip <- tempfile(fileext = ".zip")
    download.file(url, destfile = tmp_zip, mode = "wb", quiet = TRUE) #temp file path to download counties zip data to, mode = "wb" to make sure it doesn't process file as text
    
    # 2. Unzips downloaded data into a temporary folder
    tmp_dir <- tempfile()
    dir.create(tmp_dir)
    unzip(tmp_zip, exdir = tmp_dir)
    
    # 3. Finds the .shp file inside the unzipped folder and reads into R as spatial df
    shp_file <- list.files(tmp_dir, pattern = "\\.shp$", full.names = TRUE)
    roads <- st_read(shp_file, quiet = TRUE)
    
    # 4. Extract the suffix (last word of road name) and find the most common one
    result <- roads %>%
      st_drop_geometry() %>% #dropping the geometry column, don't need
      mutate(suffix = word(FULLNAME, -1)) %>% #extracting suffix, taking last word from "FULLNAME"
      distinct(FULLNAME, suffix) %>%  #remove duplicate road names within the county
      filter(
        !is.na(suffix), #removing missing values
        !suffix %in% c("N", "S", "E", "W", "NE", "NW", "SE", "SW"), #removing suffixes w/ directions
        !grepl("^[0-9]+$", suffix) #removing suffixes w/ numbers
      ) %>%
      count(suffix) %>% #counting how many roads w/ each suffix
      slice_max(n, n = 1, with_ties = FALSE) %>% #keeping most common suffix
      mutate(GEOID = geoid)
    
    # 5. Remove temporary files to free up memory (assisted w/ using Claude, it is necessary for            memory management as we are going through 3000+ county data)
    unlink(tmp_zip)
    unlink(tmp_dir, recursive = TRUE)
    rm(roads)
    
    return(result)
    
  }, error = function(e) {
    message("Failed: ", geoid, " | ", e$message) #if error occurs, returns GEOID is and reason
    return(NULL) #will move on to the next county after
  })
}
```

```{r, eval=FALSE}
# Running the function across all counties. imap_dfr() was found from the purrr as a better way to   handle compared to a regular loop, shows progress throughout the process
road_summaries <- imap_dfr(geoids, function(geoid, idx) {
  message(idx, "/", length(geoids), " | GEOID: ", geoid)
  process_county_roads(geoid)
})

# Retry any counties that failed due to server issues
failed_geoids <- geoids[!geoids %in% road_summaries$GEOID]
if (length(failed_geoids) > 0) {
  message(length(failed_geoids), "counties failed, retrying...")
  retry_summaries <- imap_dfr(failed_geoids, function(geoid, idx) {
    message(idx, "/", length(failed_geoids), " | GEOID: ", geoid)
    process_county_roads(geoid)
  })
}

# Combining first run through and the retry to final road_summaries df
road_summaries <- bind_rows(road_summaries, retry_summaries)
```

```{r}
# Filtering results down to exclude GEOID's that don't matter for project purposes. Wyoming is 56, so taking that and anything below as GEOID structure anything above isn't a state.
road_summaries <- road_summaries |>
  filter(as.numeric(substr(GEOID, 1, 2)) <= 56)

write.csv(road_summaries, "road_summaries.csv", row.names = FALSE)
```

Processing Decisions, Data Limitations

-   Duplicate Road Names: Roads appearing multiple times within the same county's shapefile were filtered out. Without removing these duplicates, the repeated entries would skew the suffix count. We used distinct() to count each unique road name only once per county.

-   Cardinal Directions: Road names that end in a directional suffix (N, S, NE, NW, etc.) have their direction captured as the suffix rather than the road type. These were filtered out as they are not meaningful road type suffixes.

-   Numbers: Some county and state route designations (County Road 40, State Route 14) end in a number rather than a standard suffix. These were filtered out as they do not represent a road type.

-   Missing Names: Some roads in the census data have no recorded name at all. We decided to remove these NA values before counting

-   Some Server Limitations: Some counties (< 10) consistently returned rejected requests from the Census website which caused us to not be able to download their data. We tried going directly to the website and it blocks us from being able to download it. "The requested URL was rejected. Please consult with your administrator". These counties are absent from the final dataset.

-   Suffix Extraction Method: Suffixes were extracted as the last word of the "FULLNAME" field. This is a reasonable approximation but is not perfect - roads with non-standard names may not be captured correctly
