---
title: "US Census Mapping Project"
subtitle: "Step 1: County Boundary Map"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 14,
  fig.height = 9
)
```

## Overview

This project uses US Census Bureau TIGER/Line shapefiles to map county boundaries
across the United States. In later steps, we will overlay road suffix data
(sourced from the Census TIGER roads files) to visualize which road suffix type
is most common in each county.

---

## Step 1: US County Boundary Map

### 1.1 Load Packages

```{r load-packages}
# install.packages(c("tigris", "ggplot2", "sf", "dplyr"))

library(tigris)   # Census TIGER/Line shapefiles
library(ggplot2)
library(sf)
library(dplyr)
```

### 1.2 Download County Shapefile

We use `tigris::counties()` to pull the 2023 cartographic boundary file directly
from the Census Bureau. Setting `cb = TRUE` gives us a simplified geometry
(smaller file, faster rendering) appropriate for a national-level map.
`options(tigris_use_cache = TRUE)` caches the download locally so repeated
knits don't re-hit the Census server.

```{r download-counties}
options(tigris_use_cache = TRUE)

counties_sf <- counties(cb = TRUE, resolution = "20m", year = 2023)

# Preview the shapefile attributes
glimpse(counties_sf)
```

> **Note on `GEOID`:** The `GEOID` column contains the 5-digit county FIPS code
> (e.g., `"01001"` for Autauga County, AL). This will be our join key when
> merging road suffix summaries in Step 3.

### 1.3 Filter to 50 States + DC

We remove US territories (Puerto Rico, Guam, etc.) that are present in the
Census file but outside the scope of this analysis.

```{r filter-counties}
# STATEFP codes for territories to exclude:
# 60 = American Samoa, 66 = Guam, 69 = Northern Mariana Islands,
# 72 = Puerto Rico,    78 = US Virgin Islands
counties_50 <- counties_sf %>%
  filter(!STATEFP %in% c("60", "66", "69", "72", "78"))

cat("Counties retained:", nrow(counties_50), "\n")
```

To restrict to the **contiguous 48 states only**, uncomment the line below:

```{r filter-contiguous, eval=FALSE}
# counties_48 <- counties_50 %>%
#   filter(!STATEFP %in% c("02", "15"))  # drop Alaska (02) and Hawaii (15)
```

### 1.4 Plot County Boundaries

We use `coord_sf(crs = 5070)` to reproject to **NAD83 / Conus Albers**
(EPSG:5070), the standard equal-area projection for national US maps. Without
reprojection, Alaska and Hawaii appear heavily distorted.

```{r plot-counties}
# --- Contiguous 48 States ---
counties_48 <- counties_50 %>%
  filter(!STATEFP %in% c("02", "15"))

p_48 <- ggplot(data = counties_48) +
  geom_sf(fill = "#f0ede8", color = "#9e9a94", linewidth = 0.1) +
  coord_sf(crs = 5070) +
  labs(
    title    = "Contiguous United States — County Boundaries",
    subtitle = "Source: US Census Bureau TIGER/Line 2023",
    caption  = "Projection: NAD83 / Conus Albers (EPSG:5070)"
  ) +
  theme_void(base_family = "serif") +
  theme(
    plot.title    = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 10, color = "gray50", hjust = 0.5, margin = margin(b = 8)),
    plot.caption  = element_text(size = 7, color = "gray60", hjust = 1),
    plot.margin   = margin(10, 10, 10, 10)
  )

p_48

# --- Alaska ---
counties_ak <- counties_50 %>%
  filter(STATEFP == "02")

p_ak <- ggplot(data = counties_ak) +
  geom_sf(fill = "#f0ede8", color = "#9e9a94", linewidth = 0.2) +
  coord_sf(crs = 3338) +   # NAD83 / Alaska Albers — standard for Alaska maps
  labs(
    title    = "Alaska — County (Borough) Boundaries",
    subtitle = "Source: US Census Bureau TIGER/Line 2023",
    caption  = "Projection: NAD83 / Alaska Albers (EPSG:3338)"
  ) +
  theme_void(base_family = "serif") +
  theme(
    plot.title    = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 10, color = "gray50", hjust = 0.5, margin = margin(b = 8)),
    plot.caption  = element_text(size = 7, color = "gray60", hjust = 1),
    plot.margin   = margin(10, 10, 10, 10)
  )

p_ak

# --- Hawaii ---
counties_hi <- counties_50 %>%
  filter(STATEFP == "15")

p_hi <- ggplot(data = counties_hi) +
  geom_sf(fill = "#f0ede8", color = "#9e9a94", linewidth = 0.2) +
  coord_sf(crs = 26904) +   # NAD83 / UTM zone 4N — standard for Hawaii maps
  labs(
    title    = "Hawaii — County Boundaries",
    subtitle = "Source: US Census Bureau TIGER/Line 2023",
    caption  = "Projection: NAD83 / UTM Zone 4N (EPSG:26904)"
  ) +
  theme_void(base_family = "serif") +
  theme(
    plot.title    = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 10, color = "gray50", hjust = 0.5, margin = margin(b = 8)),
    plot.caption  = element_text(size = 7, color = "gray60", hjust = 1),
    plot.margin   = margin(10, 10, 10, 10)
  )

p_hi
```

### 1.5 Save the Map

```{r save-map}
ggsave(
  filename = "county_map_48.png",
  plot     = p_48,
  width    = 14,
  height   = 9,
  dpi      = 200,
  bg       = "white"
)

ggsave(
  filename = "county_map_alaska.png",
  plot     = p_ak,
  width    = 14,
  height   = 9,
  dpi      = 200,
  bg       = "white"
)

ggsave(
  filename = "county_map_hawaii.png",
  plot     = p_hi,
  width    = 10,
  height   = 8,
  dpi      = 200,
  bg       = "white"
)

message("All three maps saved.")
```

---

## Summary

| Item | Detail |
|------|--------|
| Data source | US Census Bureau TIGER/Line 2023 |
| Shapefile type | Cartographic boundary (`cb = TRUE`), 1:20m resolution |
| Projection | NAD83 / Conus Albers (EPSG:5070) |
| Counties retained | `r nrow(counties_50)` |
| Join key for Step 3 | `GEOID` (5-digit county FIPS) |

The `counties_50` object will be carried forward into **Step 3** for merging
with the road suffix summaries produced in **Step 2**.
